Fly away Peter, come back Paul: the conservation and losses of avian non-coding RNAs

### 2013-10-21 paper/manuscript.tex
added Marco's RNA-seq validated RNAs.
Updated manuscript, scripts, fixed chicken seqIds, ...

%\title{Non-coding RNAs of bird genomes}
%\title{Wins and losses: Non-coding RNAs of bird genomes}
\title{Fly away Peter, come back Paul: the gains and losses of avian Non-coding RNAs}

@@

\begin{abstract}
Here we present the results of a large-scale bioinformatic annotation
of non-coding RNAs in 48 avian genomes. Our approach uses
probabilistic models of hand-curated families from the Rfam database
to infer conserved RNA families within each avian genome. We
supplement these annotations with predictions from the tRNA annotation
tool, tRNAscan-SE and microRNAs from miRBase. Although we show
extensive conservation of classical ncRNAs (e.g., tRNAs and rRNAs) and
more recently discovered ncRNAs (e.g., snoRNAs and miRNAs) in birds,
we also demonstrate apparent ``losses'' in several RNA families. These
include the divergence of some classical ncRNAs and the loss of
several snoRNAs and microRNAs. In contrast, we show a significant
number of lncRNAs are surprisingly well conserved between birds and
mammals including several intriguing cases where the reported
mammalian lncRNA function is not conserved in birds. These combined
results illustrate the utility of applying homology based methods for
annotating vertebrate genomes and illustrate many complex evolutionary
patterns within the avian ncRNA cohort.
\end{abstract}

@@

Non-coding RNAs (ncRNAs) are an important class of genes. However,
they pose serious research challenges, particularly for the field of
genomics. They lack the strong statistical signals associated with
protein coding genes, e.g. open reading frames, G+C content and
codon-usage biases \cite{Rivas:2000}.

Non-coding RNAs are classified into hierarchical groupings by the Rfam
database. The basic units are ``families'' which are groups of
homologous, alignable sequences; ``clans'' which are groups of
un-alignable (or functionally distinct), homologous families; and
``classes'' which are groups of clans and families with related
biological functions e.g. spliceosomal RNAs, miRNAs and snoRNAs
\cite{Griffiths-Jones:2003,Griffiths-Jones:2005,Gardner:2009,Gardner:2011a,Burge:2013}.
The major RNA families include the classical, highly conserved RNAs,
sometimes called ``molecular fossils'', such as the transfer RNAs,
ribosomal RNAs, RNA components of RNase P and the signal recognition
particle \cite{Jeffares:1998}. Other classes appear to have have
evolved more recently, e.g. the small nucleolar RNAs (snoRNAs),
microRNAs (miRNAs) and the long non-coding RNAs (lncRNAs)
\cite{Hoeppner:2012}. The publication of avian genomes provides
exciting opportunities to to explore ncRNA conservation in
unprecedented detail.

Despite promising techniques such as RNA-seq \cite{Croucher:2010},
homology based methods namely covariance models (CMs) are state of the
art for ncRNA analyses \cite{Sakakibara:1994,Eddy:1994,Nawrocki:2009}.
The CM based approach for annotating ncRNAs in genomes requires
reliable alignments and consensus secondary structures of
representative sequences of RNA families. These are used to train
probabilistic models for each family. These models can be used to
generate sequences with similar properties, score the likelihood that
a sequence is was generated by the same evolutionary processes as the
training sequences and to build alignments based upon sequence and
structural information \cite{Sakakibara:1994,Eddy:1994,Nawrocki:2009}.
The tRNAscan-SE software package uses CMs to accurately predict
transfer RNAs \cite{Lowe:1997,Chan:2009}. The Rfam database contains
thousands of curated alignments and consensus structures for diverse
classes of ncRNAs
\cite{Griffiths-Jones:2003,Griffiths-Jones:2005,Gardner:2009,Gardner:2011a,Burge:2013}.
Independent benchmarks of bioinformatic annotation tools have shown
that the CM approaches dramatically out-perform alternative methods
\cite{Freyhult:2007}.

@@

%Still needs a {\bf what have we done} paragraph!

\section*{Results}
In the following we explore the conservation patterns of the major
classes of ncRNA in further detail. The collection of ncRNA sequences
is generally biased towards model organisms
\cite{Gardner:2010,Hoeppner:2012}. We focus our analysis on the
unusual results within the avian lineages. These are either
unexpectedly well-conserved RNAs or unexpectedly poorly-conserved
RNAs. The former are RNAs we would not have expected to be conserved
between the birds and the organisms these genes were initially
identified in; Usually, this hypothesis is based upon the function of
the RNA which is not conserved in avian species. The latter are
apparent losses of RNA genes that were expected to be conserved;
Usually, this hypothesis is based upon the conservation of these RNAs
in other vertebrate species. We have three models of ``loss'' that we
use to explain the data: Firstly, these could be genuine cases of an
ancestral gene-loss along the avian lineage. Secondly, this could be a
case of ``divergence'' where a RNA gene has undergone significant
sequence and structural alterations, so much so, that our homology
search tools can no longer detect a relationship between vertebrate
exemplars and the avian varieties. Thirdly, we consider the
possibility that the available genome assemblies have independently
failed to capture these genes.

@@

The bulk of the ``unusually well conserved RNAs'' belong to the long
non-coding RNA (lncRNA) group.  The lncRNAs are a diverse group of
RNAs that have been implicated in a multitude of functional processes
\cite{Rinn:2007,Chow:2005,Guttman:2009}. These RNAs have largely been
characterised in mammalian species, particularly human and
mouse. Consequently, generally do not expect these to be conserved
outside of mammalia. Notable examples include Xist \cite{Duret:2006}
and H19 \cite{Smits:2008}.

In general Rfam cannot include the entire length of any large, spliced
RNAs. This is a limitation of the covariance-models used for the
homology-searches Rfam runs \cite{Nawrocki:2009}. Consequently, only
short, well-conserved regions with evolutionarily conserved secondary
structures are included in Rfam. By analogy to protein-domains, we
refer to these as RNA-domains \cite{Burge:2013}.

@@

The HOX cluster lncRNAs, HOTAIRM1, HOTTIP and HOXA11-AS1, are
remarkably well conserved. In the human genome the corresponding
RNA-domains are located at chr7:27135743-27245922 (hg19) and are
modelled by 15 RNA-domains in total, the % (5, 4 and 6 respectively)
lncRNAs are ordered HOTAIRM1 (5 RNA-domains) $\rightarrow$ HOXA11-AS1
(6 RNA-domains) $\rightarrow$ HOTTIP (4 RNA-domains). In the alligator
and turtle genomes five and six of the HOX RNA-domains are conserved,
each supporting an ancestral gene order of HOTAIRM1 $\rightarrow$
HOXA11-AS1 $\rightarrow$ HOTTIP. In the birds, where two or more of
the HOX cluster lncRNA RNA-domains were predicted on the same
scaffold, this gene order was preserved.
%egrep 'HOXA11-AS1|HOTAIR|HOTTIP' clans_competed/*gff | perl -lane 'if(/\/(\S+?)\-.*gff:(\S+).*\-id=(\S+);eval/ or /\/(\S+?)\-.*gff:(\S+).*Alias=(\S+);Not/){print "$1\t$2\t$3\t$F[6]"}'
%egrep 'PCA3' clans_competed/*gff | perl -lane 'if(/\/(\S+?)\-.*gff:(\S+).*\-id=(\S+);eval/ or /\/(\S+?)\-.*gff:(\S+).*Alias=(\S+);Not/){print "$1\t$2\t$3\t$F[6]"}'
%egrep 'RMST' clans_competed/*gff | perl -lane 'if(/\/(\S+?)\-.*gff:(\S+).*\-id=(\S+);eval/ or /\/(\S+?)\-.*gff:(\S+).*Alias=(\S+);Not/){print "$1\t$2\t$3\t$F[6]"}'

@@

%(MIR1251 is in the same locus)
%PCA3 RNA-domains inverted relative to human...

@@

{\bf more detail on the prostate lncRNAs?}

@@

%(neighbor of BRCA1 gene 2)
%MHM results!!!!
%% Another of the functions associated with lncRNAs is chromosomal dosage
%% compensation. In mammals this is carried out by the lncRNAs Xist and
%% Tsix \cite{}. In birds it has been proposed that the lncRNA MHM might
%% play a related role \cite{}.

%DLEU2, frequently deleted in malignancy, functions as a critical host
%gene of the cell cycle inhibitory microRNAs miR-15a and miR-16-1

%% TUMOR-SUPPRESSOR:
%% DLEU2, part of a tumor-suppressor cluster (with two microRNAs), most recent refs here:
%% http://www.ncbi.nlm.nih.gov/pubmed/19591824
%% http://www.ncbi.nlm.nih.gov/pubmed/20060366
%% SMAD5, tumor suppressor candidate:
%% http://www.ncbi.nlm.nih.gov/pubmed/10049768?dopt=Abstract

%% DEVELOPMENT:
%% HOXA11, most complete reference list here:
%% http://www.lncrnadb.org/Detail.aspx?TKeyID=171
%% Six3OS, retinal development:
%% http://www.ncbi.nlm.nih.gov/pubmed/21936910
%% ZEB2, brain development:
%% http://www.ncbi.nlm.nih.gov/pubmed/18347095?dopt=Abstract

%% MAT2A:
%% http://www.ncbi.nlm.nih.gov/pubmed/1426236%2C9703951?dopt=Abstract

%% RMST, doesn't look we know it does (also shows up in review paper at top):
%% http://www.ncbi.nlm.nih.gov/pubmed/20062813

%% ST7-OT, another candidate:
%% http://www.ncbi.nlm.nih.gov/pubmed/12213198?dopt=Abstract

%% PART1, meant to be human prostate-specific, birds don't even have a prostate:
%% http://www.ncbi.nlm.nih.gov/pubmed/10706094?dopt=Abstract
%% PCA3, prostate/androgen receptor signalling (curious there's heaps of it in male and female birds? same for PART1?):
%% http://www.ncbi.nlm.nih.gov/pubmed/?term=ferreira+and+pca3

%% Also, it would be good to check regarding the HOTAIRM/HOTTIP/HOXA11AS
%% etc lncRNA elements that their syntenic location relativ to the HOXA genes
%% (similar for SOXAS...) is correct. Mario checked for HOTAIRM and HOXA1/HOXA2
%% and this is fine, it would be good for my piece of mind though it is
%% at least OK in chicken, turkey, duck, zebrafinch. Also a word of caution
%% of the Rfam models: if you look at HOTAIRM1_2, there's basically no
%% covariation, looks pretty much like an ultraconserved element, so
%% technically we don't really know whether this is an RNA element or,
%% say, a TF binding site cluster that is functional at DNA level. Hence any
%% evidence that we have indeed a transcript would be nice to have. We
%% checked for HOTAIRM1 and it looks good in chicken. For the rest, it would be
%% nice to have seen transcription it at least one bird :)

@@

Vertrebrate Y RNAs typically form a cluster comprising four
well-defined paralog groups Y1, Y3, Y4, and Y5. In line with
\emph{Mosig:07a} we find that the Y5 paralog family is absent from all
bird genomes, while it is still present in both alligator and turtle,
see Supplemental Figure~4. Within bird, we find an the conserved
Y4-Y3-Y1 cluster. Apparently, broken-up clusters are in most cases
consistent with breaks in the available sequence assemblies. In
several genomes we observe one or a few additional Y RNA homologs
unlinked to the canonical Y RNA cluster. These sequences can be
identified unambiguously as derived members of one of the three
ancestral paralog groups, they almost always fit less well to the
consensus (as measured by the CM bit score of paralog group specific
covariance models) than the paralog linked to cluster, and there is no
indication that any of these additional copies is evolutionarily
conserved over longer time scales. We therefore suggest that most or
all of these interspersed copies are in fact pseudogenes.

\subsection*{MicroRNA Clusters}

@@

The microRNA family let-7 is the most abundant microRNA family with 14
paralogs in human. These genes also localize in 7 genomic clusters,
together with mir-100 and mir-125 miRNA families (see previous study on the
evolution of the let-7 miRNA cluster in \cite{Hertel:2012}). In Sauropsids
we observed that cluster A - that is strongly conserved in vertebrates has
been completely lost in the avian lineage. Another obious lost in birds is
cluster F, containing two let-7 microRNA paralogs. Cluster H, on the other
hand has been retained in all oviparous animals and completely lost later,
after the split of Theria. See Supplemental Figure~7 for details.

@@

>>>>

%\title{Non-coding RNAs of bird genomes}
%\title{Wins and losses: Non-coding RNAs of bird genomes}
\title{Fly away Peter, come back Paul: the conservation and losses of avian non-coding RNAs}

@@

\begin{abstract}
Here we present the results of a large-scale bioinformatic annotation
of non-coding RNAs in 48 avian genomes. Our approach uses
probabilistic models of hand-curated families from the Rfam database
to infer conserved RNA families within each avian genome. We
supplement these annotations with predictions from the tRNA annotation
tool, tRNAscan-SE and microRNAs from miRBase.  Although we show a
significant number of lncRNAs are surprisingly well conserved between
birds and mammals including several intriguing cases where the
reported mammalian lncRNA function is not conserved in birds.  In
contrast, we show extensive conservation of classical ncRNAs (e.g.,
tRNAs and rRNAs) and more recently discovered ncRNAs (e.g., snoRNAs
and miRNAs) in birds, we also demonstrate apparent ``losses'' in
several RNA families. These include the divergence of some classical
ncRNAs and the loss of several snoRNAs and microRNAs.  These combined
results illustrate the utility of applying homology based methods for
annotating vertebrate genomes and illustrate many complex evolutionary
patterns within the avian ncRNA cohort.
\end{abstract}

@@

Non-coding RNAs (ncRNAs) are an important class of genes. However,
they pose serious research challenges, particularly for the field of
genomics. For example, they lack the strong statistical signals
associated with protein coding genes, e.g. open reading frames, G+C
content and codon-usage biases \cite{Rivas:2000}.

Non-coding RNAs are classified into hierarchical groupings by the Rfam
database. The basic units are ``families'' which are groups of
homologous, alignable sequences; ``clans'' which are groups of
un-alignable (or functionally distinct), homologous families; and
``classes'' which are groups of clans and families with related
biological functions e.g. spliceosomal RNAs, miRNAs and snoRNAs
\cite{Griffiths-Jones:2003,Griffiths-Jones:2005,Gardner:2009,Gardner:2011a,Burge:2013}.
The major RNA families include the classical, highly conserved RNAs,
sometimes called ``molecular fossils'', such as the transfer RNAs,
ribosomal RNAs, RNA components of RNase P and the signal recognition
particle \cite{Jeffares:1998}. Other classes appear to have have
evolved more recently, e.g. the small nucleolar RNAs (snoRNAs),
microRNAs (miRNAs) and the long non-coding RNAs (lncRNAs)
\cite{Hoeppner:2012}. The publication of 48 avian genomes, including
the previously published chicken
\cite{International_Chicken_Genome_Sequencing_Consortium:2004}, zebra
finch \cite{Warren:2010} and turkey \cite{Dalloul:2010} with the newly
published 45 avian genomes \cite{}, provides exciting opportunities to
to explore ncRNA conservation in unprecedented detail.

Despite promising techniques such as RNA-seq \cite{Croucher:2010},
homology based methods namely covariance models (CMs) remain state of the
art for ncRNA analyses \cite{Sakakibara:1994,Eddy:1994,Nawrocki:2009}.
The CM based approach for annotating ncRNAs in genomes requires
reliable alignments and consensus secondary structures of
representative sequences of RNA families. These are used to train
probabilistic models for each family. These models can be used to
generate sequences with similar properties, score the likelihood that
a sequence is was generated by the same evolutionary processes as the
training sequences and to build alignments based upon sequence and
structural information \cite{Sakakibara:1994,Eddy:1994,Nawrocki:2009}.
The tRNAscan-SE software package uses CMs to accurately predict
transfer RNAs \cite{Lowe:1997,Chan:2009}. The Rfam database contains
thousands of curated alignments and consensus structures for diverse
classes of ncRNAs
\cite{Griffiths-Jones:2003,Griffiths-Jones:2005,Gardner:2009,Gardner:2011a,Burge:2013}.
Independent benchmarks of bioinformatic annotation tools have shown
that the CM approaches dramatically out-perform alternative methods
\cite{Freyhult:2007}.

@@

In the following we explore the conservation patterns of the major
classes of avian ncRNA in further detail. The collection of ncRNA
sequences is generally biased towards model organisms
\cite{Gardner:2010,Hoeppner:2012}. We focus our report on the unusual
results within the avian lineages. These are either unexpectedly
well-conserved RNAs or unexpectedly poorly-conserved RNAs. The former
are RNAs we would not have expected to be conserved between the birds
and the organisms these genes were initially identified in; Usually,
this hypothesis is based upon the function of the RNA which is not
conserved in avian species. The latter are apparent losses of RNA
genes that were expected to be conserved; Usually, this hypothesis is
based upon the conservation of these RNAs in other vertebrate species.
We use three models of ``loss'' that explain the data: Firstly, these
could be genuine cases of an ancestral gene-loss along the avian
lineage. Secondly, this could be a case of ``divergence'' where a RNA
gene has undergone significant sequence and structural alterations, so
much so, that our homology search tools can no longer detect a
relationship between vertebrate exemplars and the avian
varieties. Thirdly, we consider the possibility that the available
genome assemblies have independently failed to capture these genes.

\section*{Results}

@@

The bulk of the ``unusually well conserved RNAs'' belong to the long
non-coding RNA (lncRNA) group.  The lncRNAs are a diverse group of
RNAs that have been implicated in a multitude of functional processes
\cite{Rinn:2007,Chow:2005,Guttman:2009}. These RNAs have largely been
characterised in mammalian species, particularly human and
mouse. Consequently, we generally do not expect these to be conserved
outside of mammalia. Notable examples include Xist \cite{Duret:2006}
and H19 \cite{Smits:2008}. Some of the main results of this group are
discussed in the comparative analysis of avian genomes article \cite{}.

In general, Rfam cannot include the entire length of any large, spliced
RNAs. This is a limitation of the covariance-models used for the
homology-searches Rfam runs \cite{Nawrocki:2009}. Consequently, only
short, well-conserved regions with evolutionarily conserved secondary
structures are included in Rfam. By analogy to protein-domains, we
refer to these as RNA-domains \cite{Burge:2013}.

@@

The HOX cluster lncRNAs, HOTAIRM1, HOTTIP and HOXA11-AS1, are
remarkably well conserved. In the human genome the corresponding
RNA-domains are located at chr7:27135743-27245922 (hg19) and are
modelled by 15 RNA-domains in total, the 
lncRNAs are ordered HOTAIRM1 (5 RNA-domains) $\rightarrow$ HOXA11-AS1
(6 RNA-domains) $\rightarrow$ HOTTIP (4 RNA-domains). In the alligator
and turtle genomes five and six of the HOX RNA-domains are conserved,
each supporting an ancestral gene order of HOTAIRM1 $\rightarrow$
HOXA11-AS1 $\rightarrow$ HOTTIP. In the birds, where two or more of
the HOX cluster lncRNA RNA-domains were predicted on the same
scaffold, this gene order was also preserved.
%egrep 'HOXA11-AS1|HOTAIR|HOTTIP' clans_competed/*gff | perl -lane 'if(/\/(\S+?)\-.*gff:(\S+).*\-id=(\S+);eval/ or /\/(\S+?)\-.*gff:(\S+).*Alias=(\S+);Not/){print "$1\t$2\t$3\t$F[6]"}'
%egrep 'PCA3' clans_competed/*gff | perl -lane 'if(/\/(\S+?)\-.*gff:(\S+).*\-id=(\S+);eval/ or /\/(\S+?)\-.*gff:(\S+).*Alias=(\S+);Not/){print "$1\t$2\t$3\t$F[6]"}'
%egrep 'RMST' clans_competed/*gff | perl -lane 'if(/\/(\S+?)\-.*gff:(\S+).*\-id=(\S+);eval/ or /\/(\S+?)\-.*gff:(\S+).*Alias=(\S+);Not/){print "$1\t$2\t$3\t$F[6]"}'

@@

@@

@@

@@

Vertebrate Y RNAs typically form a cluster comprising four
well-defined paralog groups Y1, Y3, Y4, and Y5. In line with
\cite{Mosig:07a} we find that the Y5 paralog family is absent from all
bird genomes, while it is still present in both alligator and turtle,
see Supplemental Figure~4. Within bird, we find an the conserved
Y4-Y3-Y1 cluster. Apparently, broken-up clusters are in most cases
consistent with breaks in the available sequence assemblies. In
several genomes we observe one or a few additional Y RNA homologs
unlinked to the canonical Y RNA cluster. These sequences can be
identified unambiguously as derived members of one of the three
ancestral paralog groups, they almost always fit less well to the
consensus (as measured by the CM bit score of paralog group specific
covariance models) than the paralog linked to cluster, and there is no
indication that any of these additional copies is evolutionarily
conserved over longer time scales. We therefore suggest that most or
all of these interspersed copies are in fact pseudogenes.

\subsection*{MicroRNA clusters}

@@

The microRNA family let-7 is the most abundant microRNA family with 14
paralogs in human. These genes also localize in 7 genomic clusters,
together with mir-100 and mir-125 miRNA families (see previous study on the
evolution of the let-7 miRNA cluster in \cite{Hertel:2012}). In Sauropsids
we observed that cluster A - that is strongly conserved in vertebrates has
been completely lost in the avian lineage. Another obvious loss in birds is
cluster F, containing two let-7 microRNA paralogs. Cluster H, on the other
hand has been retained in all oviparous animals and completely lost later,
after the split of Theria. See Supplemental Figure~7 for details.

@@

In this work we have summarised the main results after carefully
annotating non-coding RNAs in genome sequences. We have identified
previously unrecognised conservation of ncRNAs in avian genomes as
well as some surprising ``losses'' of otherwise well conserved
ncRNAs. We suspect many of these losses are due to a combination of
limitations in the homology search tools that we use for annotation
and the ability of ncRNAs to tolerate large amounts of sequence
variation while remaining functional, rather than \emph{bona fide}
gene loss. In some cases these losses could be due to missing data
from the genome assemblies, but this unlikely to be the case for
multiple independant assemblies.

These results indicate we are still in the very early phases of
determining the functions of many RNA families. This is illustrated by
the fact that the reported functions of some ncRNAs are
mammal-specific, yet some are also found in bird genomes.
